Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-01-20 Manuscript/manuscript.Rmd
Edits through introduction

Existing methods for predicting associations between genomic elements tend to ignore key characteristics of the genomic data. Specifically, the number of TAD boundaries is much less than the number of other genomic regions, leading to heavily imbalanced classes. Furthermore, most methods utilize direct overlap as a means to quantify the association, while distance, the measure of spatial relationships, remains unaccounted for. Consequently, distances on a genomic scale vary widely, leaving uncertainty how the heavily right-tailed distribution of distance measures will affect the model’s performance.

@@

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect several epigenetic functions [@lieberman2009comprehensive]. Analyses have shown that the genome is tightly compacted into distinct compartments.  There exist regions within these compartments that have been shown to be highly conserved and self-interacting, and termed topologically associating domains (TADs) [@dixon2012topological]. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. This suggests that the boundaries of TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes. Furthermore, changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

The mechanisms underlying the formation of TADs are a complex and active area of research. Recently, it has been discovered that insulator sequences have a primary role in orchestrating the topological arrangements of higher-order chromatin architecture [@phillips2013chromatin]. Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long-range chromatin contacts [@zuin2014cohesin]. Other regulatory elements like particular histone modifications such as H3K36 and H3K27 trimethylation, and transcription factors such as ZNF274 and YY1 have been more frequently observed at TAD boundaries than in other regions [@rao20143d]. These have been found to be associated with more open chromatin which alter the accessibility of genes for transcription. The distinct patterns of different proteins and functional elements point toward the opportunity of using computational approaches in identifying which epigenomic features are most predictive of the development of TAD boundaries, allowing us to better understand what leads to their formation.

Current methods rely on the use of classification algorithms to identify which functional genomic elements are best able to discriminate between regions that include TAD boundaries and those that do not [@mourad2016computational; @hong2017computational; @huang2015predicting]. Most methods differ on the techniques used to identify genomic regions that include TAD boundaries, as well as the models and tools used to evaluate the predictive capability of said models. Moreover, many methods ignore key characteristics of domain data that can impair and bias results.

Most notably is the introduction of heavily imbalanced classes. Due to the sparcity of TADs throughout the genome, most defined regions will not contain a TAD boundary. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, this will contribute to small imbalance ratios (defined as the ratio of minority to majority classes) [@Orriols-Puig2008]. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models, especially in the case of imbalanced data.

Commonly used measures of model performance for binary classifiers include the area under the receiver operating curve (AUROC). The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a defined genomic region containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data in the area under the precision-recall curve (AUPRC). It has been shown that the use of AUPRC with imbalanced datasets yields more accurate and intuitive interpretation of model performance when compared to AUROC [@saito2015precision]. Likewise, metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data. In each case, the weight of the measure is carried primarily by the correct classification of the majority class. Better measures to consider are the F1-Score and Matthew's Correlation Coefficient (MCC) [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications. In either case, all 4 quadrants of the confusion matrix are used, making for more balanced measures in the case of imbalanced data [@chicco2017ten].

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exist several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes. Many such solutions have been developed in a variety of fields related to machine learning [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions mostly constitute as resampling techniques, in which random sampling is used to either replicate or removed data points in order to create a more balanced dataset.  It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study we focus our attention toward data level solutions so that all results may be classifier independent, and thus, more easily comparable.

Furthermore, there are other aspects of TAD boundary modelling that current methods have failed to consider. Most notably is the concept of distance. As a feature engineering step, the use of distances (in base pairs) from genomic bins to regions of functional genomic elements has largely been ignored. However, these distance characteristics have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries. Additionally, model performance is heavily contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. It is often unclear from authors of previous studies what resolution contact matrices were used when retrieving TAD boundaries, and how other resolutions would affect their results. A larger resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances.

In this study we have developed an ensemble framework that aims to address and compare a variety of data irregularities and characteristics that are associated with the prediction of TAD boundaries including class imbalance, feature engineering, and contact map resolution. We evaluated domain data obtained from four different resolution contact matrices for the GM12878 cell line including 10 kb, 25 kb, 50 kb, and 100 kb. For each resolution, we examined the inclusion of 3 different predictor types as the feature space for downstream modeling. We then performed a series of 4 different re-sampling techniques aimed at creating balanced classes. Random forest classification algorithms were performed to both compare each combination of the above ensemble, and investigate which functional genomic features were most associated with the formation of TAD boundaries.  Model performance was evaluated based on a total of four metrics which included test accuracy, area under the precision-recall curve (AUPRC), F1 score, and Matthew's Correlation Coefficient (MCC). This process was then repeated for X additional cell lines including K562, ......

>>>>

Existing methods for predicting associations between genomic elements tend to ignore key characteristics of genomic data. Specifically, the number of TAD boundaries is much less than the number of other genomic regions, leading to heavily imbalanced classes. Furthermore, most methods utilize direct overlap as a means to quantify the association, while distance, the measure of spatial relationships, remains unaccounted for. Consequently, distances on a genomic scale vary widely, leaving uncertainty how the heavily right-tailed distribution of distance measures will affect the model’s performance.

@@

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed how the spatial organization of the human genome may affect genomic regulation [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Evidence suggests that regulatory elements and genes tend to interact more frequently within the same TAD [@symmons2014functional]. Approximately 60-80% of TADs remain stable across cell types [@Schmitt:2016aa], suggestingthat the boundaries of TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes. Furthermore, changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [??? REF]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

The mechanisms underlying the formation of TADs are a complex and active area of research. Recently, it has been discovered that insulator sequences have a primary role in orchestrating the topological arrangements of higher-order chromatin architecture [@phillips2013chromatin]. Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long-range chromatin contacts [@zuin2014cohesin]. Other regulatory elements like particular histone modifications such as H3K36 and H3K27 trimethylation, and transcription factors such as ZNF274 and YY1 have been more frequently observed at TAD boundaries than in other regions [??? Better to mention those that came up in your analysis, otherwise, it is unnecessary information] [@rao20143d]. These have been found to be associated with more open chromatin which alter the accessibility of genes for transcription. The distinct patterns of different proteins and functional elements point toward the opportunity of using computational approaches in identifying which epigenomic features are most predictive of the TAD boundaries, allowing us to better understand what leads to their formation.

Current methods rely on the use of classification algorithms to identify which functional genomic elements are best able to discriminate between regions that include TAD boundaries and those that do not [@mourad2016computational; @hong2017computational; @huang2015predicting] [??? Do not understand this sentence]. Most methods differ on the techniques used to identify genomic regions that include TAD boundaries, as well as the models and tools used to evaluate the predictive capability of said models. Moreover, many methods ignore key characteristics of domain data, such as heavily imbalanced number ot boundaries vs. other regions. In a classification setting, when one is attempting to discriminate between regions with and without TAD boundaries, this will contribute to small imbalance ratios (defined as the ratio of minority to majority classes) [@Orriols-Puig2008]. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class;@chen2013influence;@jeni2013facing]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

Commonly used measures of model performance for binary classifiers include the area under the receiver operating curve (AUROC). The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a defined genomic region containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data in the area under the precision-recall curve (AUPRC). It has been shown that the use of AUPRC with imbalanced datasets yields more accurate and intuitive interpretation of model performance when compared to AUROC [@saito2015precision]. Likewise, metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. Better measures to consider are the F1-Score and Matthew's Correlation Coefficient (MCC) [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications. Both metrics use all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exist several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions mostly constitute as resampling techniques, in which random sampling is used to either replicate or removed data points in order to create a more balanced dataset [??? REFS]. It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study we focus our attention toward data level solutions so that all results may be classifier independent, and thus, more easily comparable.

Furthermore, there are other aspects of TAD boundary modelling that current methods have failed to consider. Most notably is the concept of genomic distance. As a feature engineering step, the use of distances (in base pairs) from TAD boundaries to regions of various genomic features has largely been ignored. However, these distance characteristics have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

Additionally, model performance is heavily contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. It is often unclear from previous studies what resolution contact matrices were used when retrieving TAD boundaries, and how other resolutions would affect their results. A larger resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances.

In this study we have developed an ensemble framework for predicting TAD boundaries that addresses class imbalance problem. We evaluated Hi-C data for the GM12878 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four re-sampling techniques aimed at creating balanced classes. Random forest classification was performed to both compare each combination of the above ensemble, and investigate which functional genomic features were most associated with the TAD boundaries.  Model performance was evaluated using accuracy, area under the precision-recall curve (AUPRC), F1 score, and Matthew's Correlation Coefficient (MCC). This process was then repeated for X additional cell lines including K562, ......
